---
layout: page
title: Functions and control flow
---

# Goals

We are going to build Brownian motion simulator. The idea is simple, start from
a given value (here 0), and at each generation, add a random value drawn from a
normal distribution. It's a commonly used approximation to model ecological and
evolutionary processes. Over time, the value is going on a "random walk".

The function we are going to write won't be particularly elegant or efficient
but will illustrate key programming concepts:
- how to write functions?
- how do conditional statements work?
- how do `for` loops work?

We will also use this example to get familiar with basic plotting functions.

# Step 1: Let's write a basic function

```{r}

sim_brownian <- function() {
    seq_len(10)
}

```

The curly brackets delimit the begining and the end of the function. More
generally, the curly brackets delimit a chunk of code that is meaningful in
itself.

with an argument:

```{r}

sim_brownian <- function(ngen) {
    seq_len(ngen)
}

```

with an argument and a default value:

```{r}
sim_brownian <- function(ngen=10) {
    seq_len(ngen)
}
```

# Step 2: Let's add a `for` loop

```{r}
sim_brownian <- function(ngen=10) {
    for (j in seq_len(ngen)) {
        message("j is equal to ", j)
    }
}
```

```{r}
sim_brownian <- function(ngen=10) {

    ## create a vector to hold the results
    tmp_gen <- numeric(ngen)

    for (j in seq_len(ngen)) {
        message("j is equal to ", j)
        tmp_gen[j] <- j * 2
        message("tmp_gen[j] is equal to ", tmp_gen[j])
    }
    tmp_gen
}
```

The function `rnorm()` draws random numbers from a Normal distribution. For
instance, `rnorm(1, mean=2, sd=3)` draws 1 number from a Normal distribution
with a mean 2 (default=0) and a standard deviation of 3 (default=1).

So let's modify our loop so that the vector we return holds the previous value +
a random value drawn from a Normal distribution:

```{r}
sim_brownian <- function(ngen=10) {

    ## create a vector to hold the results
    tmp_gen <- numeric(ngen)

    for (j in seq_len(ngen)) {
        message("j is equal to ", j)
        tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1)
        message("tmp_gen[j] is equal to ", tmp_gen[j])
    }
    tmp_gen
}
```

Oops... it's not working. Why?

# Step 3: Let's fix the problem

We need to tell our function that it needs to treat the first element of our
vector differently that the rest of it. That's where conditional statements come
to the rescue.

In plain English, we would say, "if this is the first element, give it the value
0, else give it the previous value + a value drawn from a Normal
distribution".

In R these are written like this:

```
if (a condition) {
	what to do if the condition is true
} else {
	what to do if the condition is false
}
```

let's modify our function:

```{r}
sim_brownian <- function(ngen=10) {

    ## create a vector to hold the results
    tmp_gen <- numeric(ngen)

    for (j in seq_len(ngen)) {
        message("j is equal to ", j)
        if (j == 1) {
            tmp_gen[j] <- 0
        } else {
            tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1)
        }
        message("tmp_gen[j] is equal to ", tmp_gen[j])
    }
    tmp_gen
}
```

## Challenge

Modify the function so that the user can specify the standard deviation of the
Normal distribution from which the random number is drawn from.

# Step 4: Let's add some replicates

Because this is an illustration of a random process, if we want to understand
and visualize the impact of changing the standard deviation on the simulations,
we need to generate replicates under the same conditions.

To do so, we are going to use nested loops. We are going to use a first loop to
create our replicates, and within each replicates, another loop to simulate the
random walk (just like before).

```{r}
sim_brownian <- function(ngen=10, nrep=10, sd) {

    ## create an empty list to store each replicate
    rep_lst <- vector("list", nrep)

    for (i in seq_len(nrep)) {
        ## create a vector to hold the results
        tmp_gen <- numeric(ngen)
        for (j in seq_len(ngen)) {
            if (j == 1) {
                tmp_gen[j] <- 0
            } else {
                tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1, sd=sd)
            }
        }
        rep_lst[[i]] <- tmp_gen
    }
    rep_lst
}
```

# Step 5: Let's plot these simulations

```{r}
sim_brownian <- function(ngen=10, nrep=10, sd) {

    ## create an empty list to store each replicate
    rep_lst <- vector("list", nrep)

    for (i in seq_len(nrep)) {
        ## create a vector to hold the results
        tmp_gen <- numeric(ngen)
        for (j in seq_len(ngen)) {
            if (j == 1) {
                tmp_gen[j] <- 0
            } else {
                tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1, sd=sd)
            }
        }
        rep_lst[[i]] <- tmp_gen
    }

    plot(seq_len(ngen), rep_lst[[1]], type="l")

    rep_lst
}
```

## Challenge

1. Add another argument to your function (call it `draw.plot`) to make the drawing
of the plot optional.
1. This is nice... but it only plot the first replicate. How can we modify the
function so that all replicates are being plotted? You need to use the function `lines()`

```{r}
sim_brownian <- function(ngen=10, nrep=10, sd) {

    ## create an empty list to store each replicate
    rep_lst <- vector("list", nrep)

    for (i in seq_len(nrep)) {
        ## create a vector to hold the results
        tmp_gen <- numeric(ngen)
        for (j in seq_len(ngen)) {
            if (j == 1) {
                tmp_gen[j] <- 0
            } else {
                tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1, sd=sd)
            }
        }
        rep_lst[[i]] <- tmp_gen
    }

    plot(seq_len(ngen), rep_lst[[1]], type="l")
    lines(seq_len(ngen), rep_lst[[2]])
    rep_lst
}
```

# Step 6: Let's fix the y-axis

```{r}

sim_brownian <- function(ngen=100, nrep=10, sd=1, draw.plot=TRUE) {

    rep_lst <- vector("list", nrep)

    for (i in seq_len(nrep)) {
        tmp_gen <- numeric(ngen)
        for (j in seq_len(ngen)) {
            if (j == 1) {
                tmp_gen[j] <- 0
            } else {
                tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1, sd=sd)
            }
        }
        rep_lst[[i]] <- tmp_gen
    }

    if (draw.plot) {
        max_y <- max(sapply(rep_lst, max))
        min_y <- min(sapply(rep_lst, min))
        plot(seq_len(ngen), rep_lst[[1]], lwd=2, type="n",
             ylim=c(min_y, max_y))

        for (i in seq_len(nrep)) {
            lines(seq_len(ngen), rep_lst[[i]])
        }
    }

    rep_lst
}

```

# Step 7: Let's make it prettier

```{r}
sim_brownian <- function(ngen=100, nrep=10, sd=1, draw.plot=TRUE) {

    rep_lst <- vector("list", nrep)

    for (i in seq_len(nrep)) {
        tmp_gen <- numeric(ngen)
        for (j in seq_len(ngen)) {
            if (j == 1) {
                tmp_gen[j] <- 0
            } else {
                tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1, sd=sd)
            }
        }
        rep_lst[[i]] <- tmp_gen
    }

    if (draw.plot) {
        max_y <- max(sapply(rep_lst, max))
        min_y <- min(sapply(rep_lst, min))
        plot(seq_len(ngen), rep_lst[[1]], lwd=2, type="n",
             ylim=c(min_y, max_y))

        for (i in seq_len(nrep)) {
            lines(seq_len(ngen), rep_lst[[i]], col="gray")
        }
    }

    rep_lst
}
```

## Challenge

1. Use  `if/else` so that the last line is drawn using a different color (e.g., "red")
1. Add two arguments to the function, so the user can specify the color s/he
would like for their lines (one argument for the color of the last line
`main.col` and an argument for the other lines `bg.col`).

Let's use the `...` to specify the labels:

```{r}
sim_brownian <- function(ngen=100, nrep=10, sd=1, draw.plot=TRUE, ...) {

    rep_lst <- vector("list", nrep)

    for (i in seq_len(nrep)) {
        tmp_gen <- numeric(ngen)
        for (j in seq_len(ngen)) {
            if (j == 1) {
                tmp_gen[j] <- 0
            } else {
                tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1, sd=sd)
            }
        }
        rep_lst[[i]] <- tmp_gen
    }

    if (draw.plot) {
        max_y <- max(sapply(rep_lst, max))
        min_y <- min(sapply(rep_lst, min))
        plot(seq_len(ngen), rep_lst[[1]], lwd=2, type="n",
             ylim=c(min_y, max_y), ...)

        for (i in seq_len(nrep)) {
            lines(seq_len(ngen), rep_lst[[i]], col="gray")
        }
    }

    rep_lst
}
```

# Step 8: Let's make it a little safer

This might be unlikely to occur (maybe you are using the function as part of
your workflow and calling it from another function that has a bug in it), but
what would happen if you tried to use the function using `nrep=0`? How about the
number of generations?

```{r}
sim_brownian(nrep=0)
sim_brownian(ngen=0, nrep=1, plot=FALSE)
```

Hmm... not so clear what the issue is. Let's create an error message that is
going to be a little more informative in this case.

Also, we may want to warn the user that if less than 10 replicates are used, the
values taken are not going to be very representative of the possible range.


```{r}

sim_brownian <- function(ngen=100, nrep=10, main.col="red",
                         bg.col="gray50", sd=1, draw.plot=TRUE, ...) {

    if (nrep < 1) {
        stop("We need at least one replicate.")
    }

    if (ngen < 2) {
        stop("We need at least 2 generations.")
    }

    if (nrep < 10)  {
        warning("With less than 10 replicates, the possible range of",
                "values may not be very representative.")
    }

    rep_lst <- vector("list", nrep)

    for (i in seq_len(nrep)) {
        tmp_gen <- numeric(ngen)
        for (j in seq_len(ngen)) {
            if (j == 1) {
                tmp_gen[j] <- 0
            } else {
                tmp_gen[j] <- tmp_gen[j - 1] + rnorm(1, sd=sd)
            }
        }
        rep_lst[[i]] <- tmp_gen
    }

    if (draw.plot) {
        max_y <- max(sapply(rep_lst, max))
        min_y <- min(sapply(rep_lst, min))
        plot(seq_len(ngen), rep_lst[[1]], col=main.col, lwd=2, type="n",
             ylim=c(min_y, max_y), ...)

        for (i in seq_len(nrep)) {
            if (i == max(nrep)) {
                lines(seq_len(ngen), rep_lst[[i]], col=main.col, lwd=2)
            } else {
                lines(seq_len(ngen), rep_lst[[i]], col=bg.col)
            }
        }
    }

    return(rep_lst)
}

```

# Things we didn't cover but need to

* scoping
* `match.arg`
